A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2017-07-03 future.tex
sjs Revisions
Update on Overleaf.

This work presented a novel methodology and Python implementation of automated
emissions harmonization for IAMs. An in-depth explanation of the processes and
methods for determining the use of harmonization methods was provided in Section
\ref{sec:meths}. The \code{aneris} code base was able to satisfactorily
harmonized over 98\% of the 2860 individual trajectories that were analyzed in
Section \ref{sec:results}. Of the remaining trajectories, harmonization method
overrides were applied, and the situation in which the need for overrides arose
was discussed.

@@

There are a variety of avenues for future improvement of both the \code{aneris}
code base and underlying methodology. As with any software project, additional
users will provide use cases for more robust handling of input/output concerns
and corner cases. Further configuration parameters may also be added in the
future in order to provide overrides for all gas species in a given sector or
region. Perhaps the most fruitful investigation will involve further refinement
of the default decision tree introduced in Section \ref{sec:meths}. A key aspect
missing from the decision tree is input from models regarding whether missing
sources are the likely cause of a harmonization year discrepancy (suggesting the
use of an offset method) or instead a significant difference in emissions
factors (suggesting the use of a ratio method) \cite{rogelj_discrepancies_2011}.

>>>>

This work presented a novel methodology and Python implementation of automated
emissions harmonization for IAMs. An in-depth explanation of the processes and
methods for determining the use of harmonization methods was provided in Section
\ref{sec:meths}. The \code{aneris} code base was able to satisfactorily
harmonized over 98\% of the 2860 individual trajectories that were analyzed in
% * <maarten.vandenberg@pbl.nl> 2017-07-03T23:07:25.083Z:
% 
% > 98\%
% Do we know to what extent does this compare to other model results?
% 
% ^.
Section \ref{sec:results}. Of the remaining trajectories, harmonization method
overrides were applied, and example situations in overrides were needed
were discussed.

@@

There are a variety of avenues for future improvement of both the \code{aneris}
code base and underlying methodology. As with any software project, additional
users will provide use cases for more robust handling of input/output concerns
and corner cases. Further configuration parameters may also be added in the
future in order to provide overrides for all gas species in a given sector or
region. Perhaps the most fruitful investigation will involve further refinement
of the default decision tree introduced in Section \ref{sec:meths}. A key aspect
missing from the decision tree is input from models regarding whether missing
sources are the likely cause of a harmonization year discrepancy (suggesting the
% * <maarten.vandenberg@pbl.nl> 2017-07-03T23:15:16.273Z:
% 
% > missing
% > sources
% it might also be a difference in activity levels, which might be something different than missing sources.
% 
% ^.
use of an offset method) or instead a significant difference in emissions
% * <maarten.vandenberg@pbl.nl> 2017-07-03T23:16:40.654Z:
% 
% > emissions
% > factors (suggesting the use of a ratio method)
% knowledge of these factors might even lead to combinations of harmonization methods for a single sector/gas/region combination as they are aggregate categories.
% 
% ^.
factors (suggesting the use of a ratio method) \cite{rogelj_discrepancies_2011}.

### 2017-07-03 intro.tex
Update on Overleaf.

The emissions trajectories calculated by IAMs are critical inputs for ongoing,
worldwide scientific community efforts in the Coupled Model Intercomparison
Project Phase 6 (CMIP6) \cite{eyring_overview_2016}, which is utilizing a number
of marker SSP scenarios developed by the IAM community
\cite{oneill_scenario_2016}. These trajectories are endogenously calculated by
modeling the individual technologies and sectors that contribute towards the
emissions of different air pollutants and GHGs as well as various mitigation
technologies (e.g., NOx and SOx scrubbers on power plants). However, the
historical emissions starting points of  models can differ by
large amounts depending on the region, sector, and emissions species.

@@

The
global climate change community has recently developed a new global historical
emissions dataset for both anthropogenic  emissions (i.e., the Community
Emissions Data System (CEDS) \cite{hoesly_historical_2017} and open-burning land-use and land-use change (LULUC)
emissions \cite{van_marle_historic_2017}) which, in
% * <sfujimori112256@gmail.com> 2017-06-07T09:59:35.820Z:
% 
% > land-use change (LUC) emissions \cite{van_marle_historic_2017}
% I think this one also anthropogenic
% 
% ^.
conjunction with the SSP IAM trajectories, will be used for climate-related
modeling exercises of CMIP6.

@@

The remainder of this paper describes the methodology and implementation of the
\code{aneris} code base \cite{matthew_gidden_2017_802832}, written in the Python
programming language (detailed documentation available
\href{http://mattgidden.com/aneris/}{online}). Section \ref{sec:meths} provides
a detailed description of the underlying mathematical components of
\code{aneris} as well as the code workflow. The results of applying the
automated harmonization mechanism on two example IAM scenarios, one with
emissions growth and another with emissions mitigation, is presented in Section
\ref{sec:results}. Finally, the general effectiveness and potential future
improvements on the automated methodology is discussed in Section
\ref{sec:future}.

>>>>

The emissions trajectories calculated by IAMs are critical inputs for ongoing,
worldwide scientific community efforts in the Coupled Model Intercomparison
Project Phase 6 (CMIP6) \cite{eyring_overview_2016}, which is utilizing a number
of marker SSP scenarios developed by the IAM community
\cite{oneill_scenario_2016}. These trajectories are endogenously calculated by
modeling the individual technologies and sectors that contribute towards the
emissions of different air pollutants and GHGs as well as various mitigation
technologies (e.g., NOx and SOx scrubbers on power plants). However, the
% * <maarten.vandenberg@pbl.nl> 2017-07-03T20:34:35.330Z:
% 
% I think the mitigation options are broad, the example of the scrubbers seems too narrow, I would leave it out.
% 
% ^.
historical emissions starting points of  models can differ by
large amounts depending on the region, sector, and emissions species.

@@

The
global climate change community has recently developed a new global historical
emissions dataset for both anthropogenic  emissions (i.e., the Community
Emissions Data System (CEDS) \cite{hoesly_historical_2017} and open-burning land-use and land-use change (LULUC)
emissions \cite{van_marle_historic_2017}) which, in
% * <sfujimori112256@gmail.com> 2017-06-07T09:59:35.820Z:
% 
% > land-use change (LUC) emissions \cite{van_marle_historic_2017}
% I think this one also anthropogenic
% 
% ^ <maarten.vandenberg@pbl.nl> 2017-07-03T20:46:42.206Z:
% 
% I think so too, however I believe the GFED also contains natural emissions.
%
% ^.
conjunction with the SSP IAM trajectories, will be used for climate-related
modeling exercises of CMIP6.

@@

The remainder of this paper describes the methodology and implementation of the
\code{aneris} code base \cite{matthew_gidden_2017_802832}, written in the Python
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:12:05.654Z:
% 
% > the
% > \code{aneris} code base \cite{matthew_gidden_2017_802832
% perhaps something like: "the harmonization code base \code{aneris}", to be explicit.
% 
% ^.
programming language (detailed documentation available
\href{http://mattgidden.com/aneris/}{online}). Section \ref{sec:meths} provides
a detailed description of the underlying mathematical components of
\code{aneris} as well as the code workflow. The results of applying the
automated harmonization mechanism on two example IAM scenarios, one with
emissions growth and another with emissions mitigation, is presented in Section
\ref{sec:results}. Finally, the general effectiveness and potential future
improvements on the automated methodology is discussed in Section
\ref{sec:future}.

### 2017-07-03 meth.tex
Update on Overleaf.

IAM emission results are provided along temporal (normally half decade or
decade), spatial (i.e., model regions), gas species, and sectoral
dimensions. Each individual temporal trajectory, i.e., unique combinations of
regions ($r$), species ($g$), and sectors ($s$), must be harmonized to the
initial starting point. Given a model trajectory, $m_{r, g, s}(t)$, historical
values, $h_{r, g, s}(t)$, and initial starting point $t_i$, a harmonized
trajectory needs to be calculated. The ``goodness of fit'' of a trajectory depends on
a number of factors, most notably the faithful representation of the original,
unharmonized, trajectory. However, other important factors must be considered,
such as the representation of negative trajectories (i.e., if a trajectory
becomes negative, both the timing and total magnitude should be as close as
possible) which are of critical importance for cumulative CO2 calculations.
% * <ssmith@pnnl.gov> 2017-07-01T03:25:04.357Z:
% 
% Suggest moving this later where negative emissions are discussed. Its a bit of a tangent here.
% 
% ^ <ssmith@pnnl.gov> 2017-07-01T03:25:44.627Z.

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). The \code{aneris} code
base provides a number of variations on the classic methods, including
ratio-convergence methods shown in Equation \ref{eqs:ratio}, offset-convergence
methods shown in Equation \ref{eqs:offset}, and interpolation methods shown in
Equation \ref{eqs:interpolate}; the convergence factor, $\beta$, for associated
methods is shown first in Equation \ref{eqs:factor}. Note that region, species,
and sector indices have been dropped for clarity. Each equation is a function of
time, model trajectory, historical trajectory, harmonization time step, and a
convergence time step ($t_f$), at which point the harmonized trajectory
converges to the unharmonized trajectory. The \code{aneris} code base provides a
number of methods to choose from for each of the harmonization families. A
summary of all available methods is provided in Table \ref{tab:meths}.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper branches represent a ``yes'' response
      and lower branches represent a ``no'' response. $dH$ is defined as
      $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and decision-making
      thresholds are described below. Finally, default choices in \textit{green}
      label methods that will likely closely match unharmonized results, choices
      in \textit{yellow} will likely somewhat match unharmonized results, and
      choices in \textit{red} can be expected to have a large relative
      difference between harmonized and unharmonized results.}
  \end{center}
\end{figure}

@@

Temporal variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method.  Emissions from forest and grassland fires, for example, vary from year to year due to a combination of meteorological conditions and anthropogenic drivers. Land use emissions are modeled by IAMs using average emission factors and do not capture conditions in a specific year.  Ratio methods for such sectors can lead to unrealistic harmonization results.
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example emissions from grassland burning: for
most models, the use of grass and paturelands is endogenously modeled, and
emissions from the burning of these lands is determined by applying an emissions
factor; in general, the more grassland usage that occurs in a scenario, the
higher the emissions from the burning thereof. Therefore, consistency in
harmonization method is important because the effects are modeled similarly
across regions and species. In order to detect land-use-like emissions,
% * <ssmith@pnnl.gov> 2017-07-01T03:31:52.910Z:
% 
% > modeled with varying fidelity by IAMs, but in general proxies are used for
% > emissions trajectories. Take for example emissions from grassland burning: for
% > most models, the use of grass and paturelands is endogenously modeled, and
% > emissions from the burning of these lands is determined by applying an emissions
% > factor; in general, the more grassland usage that occurs in a scenario, the
% > higher the emissions from the burning thereof. Therefore, consistency in
% > harmonization method is important because the effects are modeled similarly
% > across regions and species. 
% 
% Not sure I follow the logic here. I added some new text that might help. Perhaps the intermediate part could be deleted.
% 
% ^ <ssmith@pnnl.gov> 2017-07-01T03:32:30.453Z.
a measure of the coefficient of variation of the first derivative of the
historical trajectory, $c_v$, is calculated as shown in Equation \ref{eqs:cov},
where $\sigma$ is the standard deviation, and $\mu$ is the mean of the
trajectory.

The value of $c_v$ is then tested against a threshold. To determine this
threshold, an analysis of the recent CEDS and LUC historical data has been
performed. Figure \ref{fig:cov} shows the distribution of LUC $c_v$s and non-LUC
$c_v$s as determined for historical data aggregated to the model regions of 5
different flagship IAMs. A threshold value of 20 has been chosen based on these
observations as it optimally divides the two distributions. Importantly, tails
of the LUC and non-LUC overlap, thus there are both false positives (~10\% of
non-LUC trajectories) and false negatives (~7\% of LUC trajectories). However,
as any regional definition is model dependent and thus any regional aggregation
is possible an automated detection mechanism is necessary.

\begin{equation}\label{eqs:cov}
    c_v =  \frac{\sigma(h^{\prime}(t))}{\mu(h^{\prime}(t))}
\end{equation}

@@

Finally, consideration is taken with respect to the relative difference between
the historic and model values in the harmonization time period. In order to
investigate the possible values that these relative differences can take, the
flagship IAMs values used in the SSP and (ongoing) CMIP6 inter-comparison
exercises are used. A distribution of these differences for all models in the
study is presented in Figure \ref{fig:dh}. Given the available data, a threshold
value of 50\% was chosen to be used as a default in \code{aneris}.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{dh.pdf}
    \caption[]{
      \label{fig:dh}
      The distribution of relative differences between model and historical
      values in the harmonization year is shown. The solid black line indicates
      the 50\% threshold value used by default in \code{aneris}.  
    }
  \end{center}
\end{figure}

>>>>

IAM emission results are provided along temporal (normally half decade or
decade), spatial (i.e., model regions), gas species, and sectoral
dimensions. Each individual temporal trajectory, i.e., unique combinations of
regions ($r$), species ($g$), and sectors ($s$), must be harmonized to the
initial starting point. Given a model trajectory, $m_{r, g, s}(t)$, historical
values, $h_{r, g, s}(t)$, and initial starting point $t_i$, a harmonized
trajectory needs to be calculated. The ``goodness of fit'' of a trajectory depends on
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:16:50.244Z:
% 
% Assessing the "goodness of fit" is not  straightforward, there may be competing factors. Perhaps we should refer back to this in the discussion. 
% 
% ^.
a number of factors, most notably the faithful representation of the original,
unharmonized, trajectory. However, other important factors must be considered,
such as the representation of negative trajectories (i.e., if a trajectory
becomes negative, both the timing and total magnitude should be as close as
possible) which are of critical importance for cumulative CO2 calculations.
% * <ssmith@pnnl.gov> 2017-07-01T03:25:04.357Z:
% 
% Suggest moving this later where negative emissions are discussed. Its a bit of a tangent here.
% 
% ^ <ssmith@pnnl.gov> 2017-07-01T03:25:44.627Z.

In previous studies \cite{meinshausen_rcp_2011,rogelj_discrepancies_2011}, two
\textit{families} of methods have been used: those that operate on the ratio of
base year values (i.e., $\frac{h(t_i)}{m(t_i)}$) and those that operate on the
offset of base year values (i.e., $h(t_i) - m(t_i)$). The \code{aneris} code
base provides a number of variations on the classic methods, including
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:27:39.375Z:
% 
% > classic methods
% This might be interpreted as that convergence have not been previously applied. Perhaps use " implementations"  instead of "variations" .
% 
% ^.
ratio-convergence methods shown in Equation \ref{eqs:ratio}, offset-convergence
methods shown in Equation \ref{eqs:offset}, and interpolation methods shown in
Equation \ref{eqs:interpolate}; the convergence factor, $\beta$, for associated
methods is shown first in Equation \ref{eqs:factor}. Note that region, species,
and sector indices have been dropped for clarity. Each equation is a function of
time, model trajectory, historical trajectory, harmonization time step, and a
convergence time step ($t_f$), at which point the harmonized trajectory
converges to the unharmonized trajectory. The \code{aneris} code base provides a
number of methods to choose from for each of the harmonization families. A
summary of all available methods is provided in Table \ref{tab:meths}.
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:34:25.905Z:
% 
% > A
% > summary
% If beta would be defined as \frac{t -t_i}{t_f - t_i}, it might better match intuition of convergence: 0 value at t = t_i and 1 at full convergence (t = t_f), with added benefit the beta parameter can also be used in the interpolation option.
% 
% ^.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.png}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:40:43.868Z:
% 
% > The default method decision tree
% yes and no might be added to the tree for improved clarity. 
% 
% ^.
      library. For all decisions, upper branches represent a ``yes'' response
      and lower branches represent a ``no'' response. $dH$ is defined as
      $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and decision-making
      thresholds are described below. Finally, default choices in \textit{green}
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:42:35.965Z:
% 
% for CoV and dH
% 
% ^.
      label methods that will likely closely match unharmonized results, choices
      in \textit{yellow} will likely somewhat match unharmonized results, and
      choices in \textit{red} can be expected to have a large relative
      difference between harmonized and unharmonized results.}
  \end{center}
\end{figure}

@@

Temporal variability of the historical trajectory is also an important characteristic
when considering the choice of harmonization method.  Emissions from forest and grassland fires, for example, vary from year to year due to a combination of meteorological conditions and anthropogenic drivers. Land use emissions are modeled by IAMs using average emission factors and do not capture conditions in a specific year.  Ratio methods for such sectors can lead to unrealistic harmonization results.
% * <maarten.vandenberg@pbl.nl> 2017-07-03T21:59:54.238Z:
% 
% > Ratio methods for such sectors can lead to unrealistic harmonization results.
% However, using for instance an offset method may also misrepresent the emission levels. Depending on the trajectory the emission factor can be subject to unintentional changes over the century.
% 
% ^.
modeled with varying fidelity by IAMs, but in general proxies are used for
emissions trajectories. Take for example emissions from grassland burning: for
most models, the use of grass and paturelands is endogenously modeled, and
emissions from the burning of these lands is determined by applying an emissions
factor; in general, the more grassland usage that occurs in a scenario, the
higher the emissions from the burning thereof. Therefore, consistency in
harmonization method is important because the effects are modeled similarly
across regions and species. In order to detect land-use-like emissions,
% * <ssmith@pnnl.gov> 2017-07-01T03:31:52.910Z:
% 
% > modeled with varying fidelity by IAMs, but in general proxies are used for
% > emissions trajectories. Take for example emissions from grassland burning: for
% > most models, the use of grass and paturelands is endogenously modeled, and
% > emissions from the burning of these lands is determined by applying an emissions
% > factor; in general, the more grassland usage that occurs in a scenario, the
% > higher the emissions from the burning thereof. Therefore, consistency in
% > harmonization method is important because the effects are modeled similarly
% > across regions and species. 
% 
% Not sure I follow the logic here. I added some new text that might help. Perhaps the intermediate part could be deleted.
% 
% ^ <ssmith@pnnl.gov> 2017-07-01T03:32:30.453Z.
a measure of the coefficient of variation of the first derivative of the
historical trajectory, $c_v$, is calculated as shown in Equation \ref{eqs:cov},
where $\sigma$ is the standard deviation, and $\mu$ is the mean of the
trajectory.

The value of $c_v$ is then tested against a threshold. To determine this
threshold, an analysis of the recent CEDS and LUC historical data has been
performed. Figure \ref{fig:cov} shows the distribution of LUC $c_v$s and non-LUC
$c_v$s as determined for historical data aggregated to the model regions of 5
different flagship IAMs. A threshold value of 20 has been chosen based on these
observations as it optimally divides the two distributions. Importantly, tails
of the LUC and non-LUC overlap, thus there are both false positives (~10\% of
non-LUC trajectories) and false negatives (~7\% of LUC trajectories). However,
% * <maarten.vandenberg@pbl.nl> 2017-07-03T22:07:21.157Z:
% 
% >  (~10\% of
% > non-LUC trajectories) and false negatives (~7\% of LUC trajectories)
% visually it seems to be that the false negatives are a larger share thatn the false positives?
% 
% ^.
as any regional definition is model dependent and thus any regional aggregation
is possible an automated detection mechanism is necessary.

\begin{equation}\label{eqs:cov}
% * <maarten.vandenberg@pbl.nl> 2017-07-03T22:08:21.230Z:
% 
% > \begin{equation}\label{eqs:cov}
% >     c_v =  \frac{\sigma(h^{\prime}(t))}{\mu(h^{\prime}(t))}
% > \end{equation}
% c_v is called a covariance parameter and a variance of derivatives, what is exactly calculated?
% 
% ^.
    c_v =  \frac{\sigma(h^{\prime}(t))}{\mu(h^{\prime}(t))}
\end{equation}

@@

Finally, consideration is taken with respect to the relative difference between
the historic and model values in the harmonization time period. In order to
investigate the possible values that these relative differences can take, the
flagship IAMs values used in the SSP and (ongoing) CMIP6 inter-comparison
exercises are used. A distribution of these differences for all models in the
study is presented in Figure \ref{fig:dh}. Given the available data, a threshold
value of 50\% was chosen to be used as a default in \code{aneris}.
% * <maarten.vandenberg@pbl.nl> 2017-07-03T22:25:47.452Z:
% 
% > 50\%
% What does this mean? I would expect something of the form low_value < dH < high_value?
% 
% ^.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{dh.pdf}
    \caption[]{
      \label{fig:dh}
      The distribution of relative differences between model and historical
      values in the harmonization year is shown. The solid black line indicates
      the 50\% threshold value used by default in \code{aneris}.  
% * <maarten.vandenberg@pbl.nl> 2017-07-03T22:14:59.644Z:
% 
% > The distribution of relative differences between model and historical
% >       values in the harmonization year is shown. The solid black line indicates
% >       the 50\% threshold value used by defaul
% the asymmetry here might be interesting, do the SSP scenarios consistently underestimate emission levels, or is this an effect of history up to 2015, while scenarios start earlier (might not be interesting for this paper).
% 
% ^.
    }
  \end{center}
\end{figure}
